Randomized trace estimation of parameter-dependent matrices applied to spectral density approximation

### 2024-05-21 paper/main.tex
finish first draft of paper

\todo{
\begin{itemize}
    \item Unify index notation
    \item Double-check proofs
    \item Add all references
    \item Run numerical experiments 
    \item Write introduction / abstract
\end{itemize}
}

\begin{abstract}
    Abstract
\end{abstract}

@@

Need for spectral density approximation (cite some literature)

Connection to trace estimation (maybe only in words, not in formulas)

For \emph{many} evaluations of parameter, i.e. try to decrease complexity with each evaluation

Our work: Analysis added to Lin Lin, also works for general matrix functions

\subsection{Notation}
\begin{itemize}
    \item $e$ and $\log$
    \item the $k$-th moment of a random variable is denoted with $\mathbb{E}^{k}[X] = \left(\mathbb{E}\left[ | X |^{k} \right] \right)^{\frac{1}{k}}$;
    \item in a Gaussian random vector and a Gaussian random matrix, all entries follow independently the standard normal distribution;
    \item Schatten norms
    \item $\mathbb{N}$ all positive whole numbers
\end{itemize}

@@

We consider parameter-dependent matrices of the form
\begin{equation}
    \mtx{A}(t) = \begin{bmatrix}
        a_{11}(t) & a_{12}(t) & \dots & a_{1n}(t) \\
        a_{21}(t) & a_{22}(t) & \dots & a_{2n}(t) \\
        \vdots & \vdots & \ddots & \vdots \\
        a_{n1}(t) & a_{n2}(t) & \dots & a_{nn}(t) \\
    \end{bmatrix}.
\end{equation}
where $t$ is a parameter taking values in the interval $[a,b]$.
Mostly analogous to constant matrix.

Trace is defined as
\begin{equation}
    \Trace(\mtx{A}(t)) = \sum_{i=1}^{n} a_{ii}(t).
\end{equation}
If we can only access the parameter-dependent matrix through matrix-vector products, we can estimate the trace with the \emph{quadratic trace estimator} \cite{}: We take $n_{\Psi} \in \mathbb{N}$ Gaussian random vectors $\vct{\psi}_1,\vct{\psi}_2 \dots, \vct{\psi}_{n_{\psi}} \in \mathbb{R}^{n}$ to form
\begin{equation}
    \Hutch{n_{\Psi}}(\mtx{A}(t)) = \frac{1}{n_{\Psi}} \sum_{j=1}^{n_{\Psi}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j.
\end{equation}
If we gather the random vectors in a matrix $\mtx{\Psi} = [\vct{\psi}_1 ~ \vct{\psi}_2 ~ \cdots ~ \vct{\psi}_{n_{\psi}}] \in \mathbb{R}^{n \times n_{\Psi}}$ we may rewrite $\Hutch{n_{\Psi}}(\mtx{A}(t)) = n_{\Psi}^{-1} \Trace( \mtx{\Psi}^{\top} \mtx{A}(t) \mtx{\Psi})$. Other choices for the distribution of the entries of the random vectors are possible \cite{}. However, our theoretical developments only hold in the Gaussian case.

Alternatively, the trace of a symmetric matrix whose rank is significantly lower than its size can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\Omega}}$ to form the \emph{Nystr\"om approximation}
\begin{equation}
    \widehat{\mtx{A}}(t) = (\mtx{A}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} (\mtx{A}(t) \mtx{\Omega})^{\top}.
\end{equation}
Then we define the trace estimator $\Nystr{n_{\Omega}}(\mtx{A}(t)) = \Trace( \widehat{\mtx{A}}(t) )$.
Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may also rewrite
\begin{equation}
    \Nystr{n_{\Omega}}(\mtx{A}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{A}(t)^2 \mtx{\Omega})).
\end{equation}

Finally, a combination of the two estimators yields an analogue for the \emph{Nyström++ estimator} \cite{persson-2022-improved-variants} for parameter-dependent settings \todo{this is super ugly. find better notation}
\begin{equation}
    \Nystrpp{n_{\Omega}}{n_{\Psi}}(\mtx{A}(t)) = \Nystr{n_{\Omega}}(\mtx{A}(t)) + \Hutch{n_{\Psi}}(\mtx{A}(t) - \widehat{\mtx{A}}(t)).
\end{equation}
It is understood that $\widehat{\mtx{A}}$ is computed with the same sketching matrix as the one used for the trace estimator $\Nystr{n_{\Omega}}$. We can interpret this estimator as an interpolation between the trace of the Nystr\"om approximation and the quadratic trace estimator.

\subsection{Parameter-dependent quadratic trace estimator}
\label{subsec:hutchinson}

\begin{theorem}{Parameter-dependent quadratic trace estimator}{quadratic}
    Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ continuously depend on $t \in [a, b]$. Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
    \begin{equation}
        \int_{a}^{b} \left| \Trace(\mtx{A}(t)) - \Hutch{n_{\Psi}}(\mtx{A}(t)) \right| ~\mathrm{d}t < 16 \gamma k \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F  ~\mathrm{d}t.
    \end{equation}
\end{theorem}

@@

    To extend the bound to the $n_{\Psi}$-query quadratic trace estimator, we use a technique from the proof of \cite[Proof of Theorem 1]{cortinovis-2022-randomized-trace}. Let
    \begin{equation}
        \mtx{B}(t)
        = \frac{1}{n_{\Psi}} \begin{pmatrix}
            \mtx{A}(t) & & \\
            & \ddots & \\
            & & \mtx{A}(t)
        \end{pmatrix}
        \qquad \text{and} \qquad
        \vct{\psi} = \begin{pmatrix}
            \vct{\psi}_1 \\
            \vdots \\
            \vct{\psi}_{n_{\Psi}}
        \end{pmatrix}
    \end{equation}
    where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\Psi}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors. Then $\lVert \mtx{B}(t) \rVert _F = \frac{1}{\sqrt{n_{\Psi}}} \lVert \mtx{A}(t) \rVert _F$ and
    \begin{equation}
        r(t) = \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Trace(\mtx{A}(t)).
    \end{equation}
    Hence, we conclude from \refequ{equ:quadratic-trace-onequery-uniform} that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t
        < 16 k \gamma \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t.
    \end{equation}

    %Setting $\delta = \gamma^{-2 k}$ (i.e. $\gamma = \delta^{-\frac{1}{2k}}$) and choosing $k = \lceil \log(\delta^{-1}) \rceil$ we get that for all $\delta \in (0, e^{-\frac{1}{2}})$
    %\begin{equation}
    %    k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\frac{1}{2\lceil \log(\delta^{-1}) \rceil}}
    %    = \lceil \log(\delta^{-1}) \rceil e^{\frac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
    %    \leq 2 \log(\delta^{-1}) e^{\frac{1}{2}}
    %\end{equation}
    %
    %Consequently, with probability $1 - \delta$ we have
    %\begin{equation}
    %    \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\frac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t
    %\end{equation}

\subsection{Parameter-dependent Nystr\"om approximation}
\label{subsec:nystrom}

@@

We define the matrix partitions
\begin{equation}
    \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+2ex+\baselineskip}
    \mtx{U}(t) = \begin{bmatrix}
        \smash{\underbrace{\mtx{U}_1(t)}_{n \times k}} & \smash{\underbrace{\mtx{U}_2(t)}_{n \times (n-k)}}
    \end{bmatrix}
    \qquad 
    \mtx{\Lambda}(t) =
    \begin{bmatrix}
        \smash{\overbrace{\mtx{\Lambda}_1(t)}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2(t)}_{(n-k) \times (n-k)}}
    \end{bmatrix}
\end{equation}
and let $\mtx{\Omega}_1(t) = \mtx{U}_1(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times 2k}$ and $\mtx{\Omega}_2(t) = \mtx{U}_2(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times 2k}$ with $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$.

\begin{lemma}{Moment tail bound}{ose}
    A symmetric matrix $\mtx{A} \in \mathbb{R}^{l \times l}$ and stochastically independent Gaussian random matrices $\mtx{\Omega}_1 \in \mathbb{R}^{k \times 2k}, \mtx{\Omega}_2 \in \mathbb{R}^{l \times 2k}$ satisfy for all $k \in \mathbb{N}$ and a constant $c > 0$
    \begin{equation}
        \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] \leq c \left( \lVert \mtx{A} \rVert _2 + \frac{\lVert \mtx{A} \rVert _F}{\sqrt{k}} \right).
    \end{equation}
\end{lemma}

@@

\begin{lemma}{Parameter-dependent Nystrom approximation}{nystrom}
    $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ symmetric positive semi-definite and continuous in $t \in [a, b]$. $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\Omega}}$ standard Gaussian with \textcolor{red}{$n_{\Omega} \geq 8 \log(\delta^{-1})$}. Then with probability at least $1 - \delta$
    \begin{equation}
        \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F ~\mathrm{d}t
        \leq C_{\Omega}\frac{1}{\sqrt{n_{\Omega}}} \int_{a}^{b} \Trace(\mtx{A}(t)) ~\mathrm{d}t 
    \end{equation}
\end{lemma}

\todo{proof idea}

\begin{proof}
    Omitting the parameter dependence, we can apply \cite[Theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound the integrand with
    \begin{equation}
        \lVert \mtx{A} - \widehat{\mtx{A}} \rVert _F 
        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert (\mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F.
    \end{equation}

    As in the proof of \cite[Lemma 3]{meyer-2021-hutch-optimal}, the first term can be bonud with
    \begin{equation}
        \lVert \mtx{\Lambda}_2 \rVert _F
        = \sqrt{\sum_{i=k+1}^{n} \lambda_i^2}
        \leq \sqrt{ \lambda_{k+1} \sum_{i=k+1}^{n} \lambda_i}
        \leq \sqrt{ \frac{\Trace(\mtx{A})}{k} \Trace(\mtx{A})}
        \leq \frac{1}{\sqrt{k}} \Trace(\mtx{A}).
    \end{equation}

>>>>

%\todo{
%\begin{itemize}
%    \item Unify index notation
%    \item Double-check proofs
%    \item Add all references
%    \item Run numerical experiments 
%    \item Write introduction, abstract, and discussion
%\end{itemize}
%}

\begin{abstract}
    \todo{Write abstract.}
\end{abstract}

@@

\begin{itemize}
    \item Definition of parameter-dependent matrix ($\mtx{B}(t) \in \mathbb{R}^{n \times n}$, every entry depends continuously on $t \in [a,b]$ \cite{kressner-2023-randomized-lowrank})
    \item Trace estimation through mat-vecs (if matrix is only accessed implicitly [Hessian, Graph] or for matrix functions [partition function, spectral density, exponential integrators])
    \item Mention existing trace estimation methods for constant matrices (quadratic [Hutchinson, Girard], Hutch++ \cite{meyer-2021-hutch-optimal}, Nyström++ \cite{persson-2022-improved-variants}, XTrace [Tropp], Krylov aware [Chen]...)
\end{itemize}

\paragraph{Spectral density approximation}
\begin{itemize}
    \item Definition of spectral density ($\phi(t) = n^{-1}\sum_{i=1}^{m} \delta(t - \lambda_i)$ for matrix $\mtx{A}$ with eigenvalues $\lambda_1, \dots, \lambda_n$)
    \item Applications of spectral density approximation (electronic structure calculations \cite{lin-2017-randomized-estimation}, neural network optimization [Hessian, ...], graph processing [filtering, ...])
    \item Link to trace estimation of a matrix function (smoothing with Gaussian \cite{lin-2017-randomized-estimation} or Haydock [Haydock, Lin] kernel)
    \item Existing theory for spectral density approximation (Wasserstein-1 error [Braverman, Chen], missing $1/\varepsilon$ guarantees)
\end{itemize}

\paragraph{Contribution}
\begin{itemize}
    \item Development of a unified family of estimators for many evaluations of parameter, i.e. try to keep complexity with each evaluation as low as possible
    \item Extends \cite{kressner-2023-randomized-lowrank} to trace estimation
    \item Provides theoretical guarantees for methods in \cite{lin-2017-randomized-estimation}
\end{itemize}

\paragraph{Notation}
We follow the notation and conventions which are used in contemporary literature in this field. Further,
\begin{itemize}
    \item we denote with $\mathbb{N}$ all positive whole numbers and with $\mathbb{R}$ all real numbers;
    \item $e$ denotes Euler's number and $\log$ is the corresponding logartihm;
    \item the ceiling function $\lceil x \rceil$ of a real number $x$ returns the greatest integer larger than or equal to $x$;
    \item scalars are represented by lower case Greek or Latin letters ($s$, $\varepsilon$, \dots),
    vectors are additionally printed in bold ($\vct{v}$, $\vct{\psi}$, \dots),
    and matrices are additionally capitalized ($\mtx{A}$, $\mtx{\Omega}$, \dots);
    \item the expected value of a random variable $x$ is denoted with $\mathbb{E}[x]$. The $k$-th moment of a random variable $x$ is defined as $\mathbb{E}^{k}[x] = \left(\mathbb{E}\left[ | x |^{k} \right] \right)^{\frac{1}{k}}$;
    \item the identity matrix $\mtx{I}_n = \diag(1, \dots, 1) \in \mathbb{R}^{n \times n}$
    carries ones on its diagonal and zeros everywhere else. The zero matrix
    $\mtx{0}_{n \times m} \in \mathbb{R}^{n \times m}$ consists of only zero entries;
    \item in a Gaussian random vector and a Gaussian random matrix, all entries are independently sampled from the standard normal distribution;
    \item matrix norms are denoted with $\lVert \cdot \rVert _{\cdot}$. For symmetric matrices, the $p$-Schatten norm is
    $\lVert \mtx{A} \rVert _{(p)} = (\sum_{i=1}^n |\lambda_i|^p)^{1/p}$. We refer to the spectral norm as $\lVert \mtx{A} \rVert _2 = \lVert \mtx{A} \rVert _{(\infty)} = \max_{i=1,\dots,n}|\lambda_i|$, the nuclear norm as $\lVert \mtx{A} \rVert _{\ast} = \lVert \mtx{A} \rVert _{(1)} = \sum_{i=1}^{n} |\lambda_i|$, and the Frobenius norm as $\lVert \mtx{A} \rVert _F = \lVert \mtx{A} \rVert _{(2)} = (\sum_{i=1}^{n} |\lambda_i|^2)^{\frac{1}{2}}$.
\end{itemize}

@@

We consider parameter-dependent matrices of the form
\begin{equation}
    \mtx{A}(t) = \begin{bmatrix}
        a_{11}(t) & a_{12}(t) & \dots & a_{1n}(t) \\
        a_{21}(t) & a_{22}(t) & \dots & a_{2n}(t) \\
        \vdots & \vdots & \ddots & \vdots \\
        a_{n1}(t) & a_{n2}(t) & \dots & a_{nn}(t) \\
    \end{bmatrix} \in \mathbb{R}^{n \times n}
\end{equation}
where $a_{ij}(t)$ are functions depending continuously on the parameter $t$ which takes values in the interval $[a,b]$. The trace of such a matrix is defined as
\begin{equation}
    \Trace(\mtx{A}(t)) = \sum_{i=1}^{n} a_{ii}(t).
\end{equation}

\paragraph{Quadratic trace estimator} If we can only access the parameter-dependent matrix through matrix-vector products, we can estimate the trace with the quadratic trace estimator \cite{}: We take $n_{\Psi} \in \mathbb{N}$ Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\psi}} \in \mathbb{R}^{n}$ to form
\begin{equation}
    \Hutch{n_{\Psi}}(\mtx{A}(t)) = \frac{1}{n_{\Psi}} \sum_{j=1}^{n_{\Psi}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j.
    \label{equ:quadratic-trace-estimator}
\end{equation}
If we gather the random vectors in a matrix $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\psi}}] \in \mathbb{R}^{n \times n_{\Psi}}$ we may rewrite $\Hutch{n_{\Psi}}(\mtx{A}(t)) = n_{\Psi}^{-1} \Trace( \mtx{\Psi}^{\top} \mtx{A}(t) \mtx{\Psi})$. Other choices for the distribution of the entries of the random vectors are possible \cite{}. However, our theoretical developments only hold in the Gaussian case.

\paragraph{Nyström trace approximation} Alternatively, the trace of a symmetric matrix whose rank is significantly lower than its size can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\Omega}}$ to form the Nyström approximation
\begin{equation}
    \widehat{\mtx{A}}(t) = (\mtx{A}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} (\mtx{A}(t) \mtx{\Omega})^{\top}.
\end{equation}
Then we define the trace estimator $\Nystr{n_{\Omega}}(\mtx{A}(t)) = \Trace( \widehat{\mtx{A}}(t) )$.
Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may also rewrite
\begin{equation}
    \Nystr{n_{\Omega}}(\mtx{A}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{A}(t)^2 \mtx{\Omega})).
    \label{equ:nystrom-trace-estimator}
\end{equation}

\paragraph{Nyström++ estimator} Finally, a combination of the two estimators yields an analogue for the Nyström++ estimator \cite{persson-2022-improved-variants} in the parameter-dependent setting
\begin{equation}
    \Nystrpp{n_{\Omega}}{n_{\Psi}}(\mtx{A}(t)) = \Nystr{n_{\Omega}}(\mtx{A}(t)) + \Hutch{n_{\Psi}}(\mtx{A}(t) - \widehat{\mtx{A}}(t)).
    \label{equ:nystrompp-trace-estimator}
\end{equation}
It is understood that $\widehat{\mtx{A}}$ is computed with the same sketching matrix as the one used for the trace estimator $\Nystr{n_{\Omega}}$. We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the quadratic trace estimator.

\paragraph{$L^{1}$ norm}
\todo{Give definition of $L^1$ norm and reason for choosing it}

\subsection{Parameter-dependent quadratic trace estimator}
\label{subsec:quadratic}

We first consider \refequ{equ:quadratic-trace-estimator}.
\todo{Discuss theory for constant matrix
\begin{equation}
    \lVert f \rVert _1 = \int_{a}^{b} | f(t) | \mathrm{d}t
\end{equation}
}

\begin{theorem}{Parameter-dependent quadratic trace estimator}{quadratic}
    Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ continuously depend on $t \in [a, b]$. Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
    \begin{equation}
        \int_{a}^{b} \left| \Trace(\mtx{A}(t)) - \Hutch{n_{\Psi}}(\mtx{A}(t)) \right| ~\mathrm{d}t < 16 \gamma k \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F  ~\mathrm{d}t.
    \end{equation}
    In particular, if we choose $n_{\Psi} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{\frac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Hutch{n_{\Psi}}(\mtx{A}(t)) | ~\mathrm{d}t < \varepsilon \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t$.
\end{theorem}

@@

    To extend the bound to the $n_{\Psi}$-query quadratic trace estimator, we use a technique from the proof of \cite[Proof of Theorem 1]{cortinovis-2022-randomized-trace}. Let
    \begin{equation}
        \mtx{B}(t)
        = \frac{1}{n_{\Psi}} \begin{pmatrix}
            \mtx{A}(t) & & \\
            & \ddots & \\
            & & \mtx{A}(t)
        \end{pmatrix}
        \quad \text{and} \quad
        \vct{\psi} = \begin{pmatrix}
            \vct{\psi}_1 \\
            \vdots \\
            \vct{\psi}_{n_{\Psi}}
        \end{pmatrix}
    \end{equation}
    where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\Psi}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors. Then $\lVert \mtx{B}(t) \rVert _F = \frac{1}{\sqrt{n_{\Psi}}} \lVert \mtx{A}(t) \rVert _F$ and
    \begin{equation}
        r(t) = \frac{1}{n_{\Psi}} \sum_{j=1}^{n_{\Psi}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j - \Trace(\mtx{A}(t)).
    \end{equation}
    Hence, we conclude from \refequ{equ:quadratic-trace-onequery-uniform} that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{j=1}^{n_{\Psi}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t
        < 16 k \gamma \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t.
    \end{equation}

    Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) \rceil$ we get that for all $\delta \in (0, e^{-\frac{1}{2}})$
    \begin{equation}
        k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\frac{1}{2\lceil \log(\delta^{-1}) \rceil}}
        = \lceil \log(\delta^{-1}) \rceil e^{\frac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
        \leq 2 \log(\delta^{-1}) e^{\frac{1}{2}},
    \end{equation}
    from which follows the second part of the theorem.
    %Consequently, with probability $1 - \delta$ we have
    %\begin{equation}
    %    \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\frac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t
    %\end{equation}
\end{proof}

\todo{Discuss $1/\varepsilon^2$ drawback}

\subsection{Parameter-dependent Nyström approximation}
\label{subsec:nystrom}

Next we consider \refequ{equ:nystrom-trace-estimator}. \todo{Motivate Nyström approximation if matrix is low-rank}

@@

We define the matrix partitions
\begin{equation}
    \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+2ex+\baselineskip}
    \mtx{U}(t) = \begin{bmatrix}
        \smash{\underbrace{\mtx{U}_1(t)}_{n \times k}} & \smash{\underbrace{\mtx{U}_2(t)}_{n \times (n-k)}}
    \end{bmatrix}
    \quad \text{and} \quad
    \mtx{\Lambda}(t) =
    \begin{bmatrix}
        \smash{\overbrace{\mtx{\Lambda}_1(t)}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2(t)}_{(n-k) \times (n-k)}}
    \end{bmatrix}
\end{equation}
and let $\mtx{\Omega}_1(t) = \mtx{U}_1(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times 2k}$ and $\mtx{\Omega}_2(t) = \mtx{U}_2(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times 2k}$ with $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$.

\begin{lemma}{Moments of tail bound}{ose}
    A symmetric matrix $\mtx{A} \in \mathbb{R}^{l \times l}$ and stochastically independent Gaussian random matrices $\mtx{\Omega}_1 \in \mathbb{R}^{k \times 2k}, \mtx{\Omega}_2 \in \mathbb{R}^{l \times 2k}$ satisfy for all $k \in \mathbb{N}$ and a constant $c > 0$
    \begin{equation}
        \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] \leq c \left( \lVert \mtx{A} \rVert _2 + \frac{\lVert \mtx{A} \rVert _F}{\sqrt{k}} \right).
    \end{equation}
\end{lemma}

@@

\begin{lemma}{Parameter-dependent Nystrom approximation}{nystrom}
    Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$. Then its Nyström approximation $\widehat{\mtx{A}}(t)$ with Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\Omega}}$ for even $n_{\Omega} \in \mathbb{N}$ and $\gamma \geq 1$ verifies with probability at least $1 - \gamma^{-\frac{n_{\Omega}}{4}}$
    \begin{equation}
        \int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F~\mathrm{d}t \leq \gamma \sqrt{2} \cdot \frac{1 + 2c}{\sqrt{n_{\Omega}}} \int_{a}^{b} \Trace(\mtx{A}(t))~\mathrm{d}t.
    \end{equation}
    for a universal constant $c > 0$. In particular, if we choose $n_{\Omega} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})), n_{\Omega} \geq 2$, then with probability at least $1-\delta$ for $\delta \in (0, e^{-e^{4 \varepsilon^2}})$ and $\varepsilon > 0$ we have $\int_{a}^{b} \lVert \mtx{A}(t) - \widehat{\mtx{A}}(t) \rVert _F~\mathrm{d}t  < \varepsilon \int_{a}^{b} \Trace(A(t))~\mathrm{d}t$.
\end{lemma}

\todo{proof idea (structural bound, then higher order moment bound to apply Markov's inequality)}

\begin{proof}
    We let $n_{\Omega} = 2k$ for some $k \in \mathbb{N}$, i.e. $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$. After temporarily omitting the parameter dependence, we can apply \cite[Theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound the integrand with \todo{maybe explain where $\mtx{\Omega}_i$ come from...}
    \begin{equation}
        \lVert \mtx{A} - \widehat{\mtx{A}} \rVert _F 
        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert (\mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F.
        \label{equ:nystrom-proof-persson-bonud}
    \end{equation}

    As in the proof of \cite[Lemma 3]{meyer-2021-hutch-optimal}, the first term can be bonud with
    \begin{equation}
        \lVert \mtx{\Lambda}_2 \rVert _F
        = \sqrt{\sum_{i=k+1}^{n} \lambda_i^2}
        \leq \sqrt{ \lambda_{k+1} \sum_{i=k+1}^{n} \lambda_i}
        \leq \sqrt{ \frac{\Trace(\mtx{A})}{k} \Trace(\mtx{A})}
        \leq \frac{1}{\sqrt{k}} \Trace(\mtx{A}).
        \label{equ:nystrom-proof-frobenius-trace}
    \end{equation}

    The second term is first processed with standard matrix-norm inequalities:
    \begin{equation}
        \lVert (\mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F 
        \leq \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F
        \leq \sqrt{k} \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2.
        \label{equ:nystrom-proof-processed-tail}
    \end{equation}
    Its higher order moments are bounded by
    \begin{align}
        \mathbb{E}^{\frac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
        &= \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2 && \text{(by definition of $\mathbb{E}^{k}$)} \notag \\
        &\leq c \left( \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _F \right)^2 && \text{(\reflem{lem:ose} with $\mtx{A}=\mtx{\Lambda}_2^{\frac{1}{2}}$)} \notag \\
        &\leq c \left( \lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _2^2  + \frac{1}{k}\lVert \mtx{\Lambda}_2^{\frac{1}{2}} \rVert _F^2 \right) && \text{($(a + b)^2 \leq a^2 + b^2$)} \notag \\
        &= c \left( \lambda_{k+1} + \frac{1}{k}\Trace(\mtx{\Lambda}_2) \right) && \text{(definition of $\lVert \cdot \rVert _2$ and $\lVert \cdot \rVert _F$)} \notag \\
        &\leq \frac{2 c}{k} \Trace(\mtx{A}) && \text{($\lambda_{k+1} \leq \frac{\Trace(\mtx{A})}{k}$, $\Trace(\mtx{\Lambda}_2) \leq \Trace(\mtx{A})$)}
        \label{equ:nystrom-proof-tail-bound}
    \end{align}
    Thus, inserting \refequ{equ:nystrom-proof-tail-bound} into \refequ{equ:nystrom-proof-processed-tail} and using it along with \refequ{equ:nystrom-proof-frobenius-trace} into \refequ{equ:nystrom-proof-persson-bonud} we obtain
    \begin{equation}
        \mathbb{E}^{\frac{k}{2}} \left[\lVert \mtx{A} - \widehat{\mtx{A}} \rVert _F \right]
        \leq \frac{1 + 2c}{\sqrt{k}} \Trace(\mtx{A})
    \end{equation}
